---
title: 'Large scale biodiversity patterns'
author: Jesús N. Pinto-Ledezma
output:
  pdf_document: default
  html_document: 
    theme: readable
    toc: yes
---
# Short intro
Since the eighteenth century, broad-scale patterns of diversity called the attention of naturalists. Recognizing that tropical regions have higher species richness relative to temperate areas, Alexander von Humboldt was the first one to propose it to emerge from climatic differences **(Hawkins 2001)**. This ubiquitous pattern has since then been known as the Latitudinal Diversity Gradient (LDG) and, although the global distribution of biodiversity is indeed far more complex than a simple unidirectional gradient **(Hawkins and Diniz-Filho 2004)**, the difference in species richness between temperate and tropical regions tends to capture the most evident facet of the distribution of life on Earth: its geographic heterogeneity.

Early explanations for the LDG in the 1950s and 1960s followed von Humboldt’s tradition and focused on the strong correlations observed between diversity (i.e., species richness) and components of current environmental variation—especially combinations of temperature and precipitation **(Simpson 1964; Pianka 1966; O’Brien 2006; Hawkins et al. 2003; Brown 2012)**. These high correlations suggested a causal explanation, and spurred the development of hypotheses that aimed to identify the mechanisms affecting species distributions and hence driving geographical patterns **(e.g., Currie et al. 2004)**. Although these diversity-environment correlations suggested “pure ecological explanations” that involved population-level processes tied to dispersal and aggregation of tropical organisms, it quickly became clear that deep-time evolutionary processes should also be taken into account to explain the LDG **(Rohde 1992; Ricklefs 2004)**. In fact, as early as 1937, Theodosius Dobzhansky had proposed that diversity gradients should be explained by an interaction between ecological and evolutionary mechanisms, in which evolution would drive the dimensions of the niche—the set of biotic and abiotic factors that allow a species to exist indefinitely—that would allow different patterns of niche packing throughout environmental gradients. Today, it is consensus that the LDG should be explained not only by current climatic factors, but also by the long-term dynamics of such climatic factors and by events happening throughout the evolution of the species **(Fine 2015)**.

Today we will learn basic tools in R for visualizing species distributions, build geographical ranges, testing drivers of gradients of biodiversity under different approaches.

You will need four datasets, that will be provided for you:
  1. Species occurrence data points - live.oaks.txt
  2. Species geograhical ranges - Furnarii_ranges_geo.shp
  3. Phylogenetic tree - furnariiMCC.nex
  4. Environmental predictors - bio1.bil and bio12.bil

# Set up your data and your working directory

Set up a working directory and put the data files in that directory. Tell R that this is the directory you will be using, and read in your data:

```{r, warnings = FALSE, message = FALSE, eval = FALSE}
setwd("path/for/your/directory")
```

Install and load the following packages

```{r, eval = FALSE}
require(maptools)
require(rgdal)
require(raster)
require(sp)
require(rangeBuilder)
library(spdep)
library(ncf)
require(geiger)
require(dismo)
library(letsR)
library(rworldmap)
require(spatialreg)
require(picante)
require(ape)
```
# From point occurrences to range maps

Load species occurrences data points. We will use occurrences from Live oaks, that were obtained from iDigBio between 20 and 24 July 2018 by Jeannine Cavender-Bares. Notice that these occurrence data points were visually examined and any localities that were outside the known range of the species, or in unrealistic locations (e.g., water bodies, crop fields) were discarded. 

```{r, warnings = FALSE, message = FALSE, eval = FALSE}
oaks <- read.table("Data/OCC/live.oaks.txt", header = TRUE)
```

let's start exploring the data.
```{r, warnings = FALSE, message = FALSE, eval = FALSE}
head(oaks)

tail(oaks)
```

*What variables we have in the object oaks? How many oak species are in the dataset?*

Plot the points (x = Logitude, y = Latitude) and a world map, for reference. We need to load a data object from the maptools package

```{r, eval = FALSE, fig.keep='all'}
plot(oaks[c(2:3)], col = "blue", pch = 19)
plot(countriesCoarse, add = TRUE)
```

Let's try with a single species in this case, *Quercus virginiana*
```{r, eval = FALSE}
unique(oaks$Species)

que_vir <- subset(oaks, oaks$Species == "Quercus_virginiana")
```

```{r, eval = FALSE}
plot(que_vir$Longitude, que_vir$Latitude, pch = 15)
plot(countriesCoarse, add = TRUE)
```

Cool, right? 

# Data checking and cleaning

Check if there are any duplicated points. We will check if exist any duplicate occurrence data point, if so, then remove all duplicates.

```{r, eval = FALSE}
oaks_dups <- duplicated(oaks[, c(2:3)])
### NOTE: the function "duplicated" returns the results of a logical test (e.g. TRUE or FALSE)
# How many are duplicates?
length(which(oaks_dups == TRUE))
# How many are NOT duplicates?
length(which(oaks_dups == FALSE))
# Keep only those lines that are not duplicates
oaks_dups_row <- which(oaks_dups == TRUE)
# What's the size? That is, how many points are duplicates
length(oaks_dups_row)
# Create another object withoyt the duplicate records
oaks_nodups <- oaks[-oaks_dups_row,]
# What are the dimensions of the new object?
dim(oaks_nodups)
# Take a look at the first rows of data
head(oaks_nodups)
```

*What was the result? Are there duplicated occurrences? if so, how many?*

Lets plot the results!

```{r, eval = FALSE}
plot(oaks$Longitude, oaks$Latitude, pch = 19, col = "red", cex = 2)
points(oaks_nodups$Longitude, oaks_nodups$Latitude, pch = 16, col = "black")
```

# Range maps from point data

This part can be used to create “simple” range maps based on geometry (e.g. minimum convex polygons, etc.), without considering environmental variables (no ENMs or SDMs). Note that these range maps are geographical abstractions of the species ranges. **A species range is the area where a particular species can be found during its lifetime. Species range includes areas where individuals or communities may migrate or hibernate**

## Convex hull (minimum convex polygon)

The convex hull may be defined either as the intersection of all convex sets containing a given subset of a Euclidean space, or equivalently as the set of all convex combinations of points in the subset. Convex hulls of open sets are open, and convex hulls of compact sets are compact. Every convex set is the convex hull of its extreme points. 

This model draws a convex hull around all ’presence’ points.

```{r, eval = FALSE}
# create a polygon around the species' records
oaks_hull <- convHull(oaks_nodups[, c(2:3)])
str(oaks_hull)
```


```{r, eval = FALSE}
# Plot the created polygon
plot(oaks_hull)
points(oaks_nodups$Longitude, oaks_nodups$Latitude, pch = 16, col = "black")
```


Ok, now let's try with a single species,

```{r, eval = FALSE}
que_vir <- subset(oaks_nodups, oaks_nodups$Species == "Quercus_virginiana")
```


```{r, eval = FALSE}
que_vir_hull <- convHull(que_vir[, c(2:3)])
```

```{r, eval = FALSE}
plot(que_vir_hull)
points(que_vir$Longitude, que_vir$Latitude, pch = 16, col = "black")
```

Now lets plot all ranges.

```{r, eval = FALSE}
# Plot all live oaks
plot(oaks_hull)
points(oaks_nodups$Longitude, oaks_nodups$Latitude, pch = 16, col = "black")

# Plot only Quecus virginiana
plot(que_vir_hull, add = TRUE)
points(que_vir$Longitude, que_vir$Latitude, pch = 16, col = "red")

# Add world maps
plot(countriesCoarse, add = TRUE)
```

Ok, using simple convex hull seems not to be a good tool for live oaks, now lets try another approach. Now, we will use the dynamic alpha hull from the package rangeBuilder.

## Dynamic Alpha hull

```{r, eval = FALSE}
que_vir_alphahull <- getDynamicAlphaHull(que_vir, fraction = 0.95, 
                                         coordHeaders = c("Longitude", "Latitude"),
                                         clipToCoast = 'no')[[1]]
```

Now plot and look at the differences.

```{r, eval = FALSE}
plot(que_vir_alphahull, lwd = 2, col = "red") 
plot(que_vir_hull, add = TRUE, lwd = 2, lty = 2)
points(que_vir$Longitude, que_vir$Latitude, pch = 16, col = "green")
plot(countriesCoarse, add = TRUE, lwd = 2)
```

*Is there a difference? Please, explain the difference!*

Until here we explored how to plot, clean and build species geographical ranges using occurrences. Now will use species geographical ranges of the largest continental endemic radiation (Furnariides) to explore the geographical gradients of species diversity.

# Diversity gradients

## Prepare data and mapping

The geographical ranges correspond to the Infraorder Furnariides (Aves). This data is available thorough BirdLife International (http://datazone.birdlife.org/species/requestdis) and you can use any other group available on IUCN or BIEN (for plants in the Americas). In any case, you first need to download the polygons in shapefile format.

To load the Furnariides geographical ranges we will use the function **readOGR** from the package **rgdal**. There are other functions

```{r, eval = FALSE}
franges <- readOGR(dsn = "Data/Franges", layer = "Furnarii_ranges_geo")
```

Now explore the data inside the ranges. Notice that to access to the information, we will use **@** instead of **$**. 

```{r, eval = FALSE}
head(franges@data)
```

Let's plot a couple of species.

```{r, eval = FALSE}
fur_ruf <- subset(franges, franges$SCINAME == "Furnarius rufus")
ana_dor <- subset(franges, franges$SCINAME == "Anabazenops dorsalis")
```

```{r, eval = FALSE}
plot(fur_ruf, col = "green") # Furnarius rufus
plot(ana_dor, add = TRUE, col = "red") # Anabazenops dorsalis
# Add world maps
plot(countriesCoarse, add = TRUE)
```

*Explain the distribution of the Furnarius rufus*

## Raster of species richness

Species richness is the number of different species represented in an ecological community, landscape or region. Species richness is simply a count of species, and it does not take into account the abundances of the species or their relative abundance distributions.

Now, let's create the a map that represent the species richness of Furnariides.

First create an empty raster for the Neotropics using the extent of the furnariides ranges under a spatial resolution of 1º long-lat.

```{r, eval = FALSE}
neo_ras <- raster()
# Set the raster "extent" 
extent(neo_ras) <- extent(franges)
res(neo_ras) <- 1

neo_ras

values(neo_ras) <- 0
```

Now using the empty raster we will **rasterize** the species identities in each cell or pixel. The resulting raster will be the species richness of Furnariides across the Neotropics.

```{r, eval = FALSE}
f_sr_raster <- rasterize(franges, neo_ras, field = "SCINAME", 
                         fun = function(x,...){length(unique(na.omit(x)))})
```

Plot the raster.

```{r, eval = FALSE}
plot(f_sr_raster)
plot(countriesCoarse, add = T)
```

Let's try changing the colors.

```{r, eval = FALSE}
#change the color scale
colfuncYellows <- colorRampPalette(c("#d7191c", "#fdae61", "#ffffbf", "#abd9e9", 
                                     "#2c7bb6"))
````

```{r, eval = FALSE}
plot(f_sr_raster, col = rev(colfuncYellows(100)), axes = FALSE, box = FALSE, 
     zlim = c(minValue(f_sr_raster), maxValue(f_sr_raster)), 
     xlab = "Furnariides richness", legend.width = 2)
plot(countriesCoarse, add = T)
```

*Awesome, right?. Now, please describe the observed pattern!*

Lets try to rasterize other information from the polygon data set. We will use the information in the column **RD**, this data correspond to the numbers of nodes from the tips to the root of a phylogenetic tree or just **root distance**, thus, will use the RD to calculate the MRD metric **(mean root distance)** that measures the evolutionary derivedness of species within an assemblage **(Kerr & Currie, 1999)** and can be used to determine whether a local fauna is constituted primarily by early-diverged or by recently originated species **(Hawkins et al., 2012)**. 

```{r, eval = FALSE}
head(franges@data)

f_MRD_raster <- rasterize(franges, neo_ras, field = "RD", fun = mean)
```

```{r, eval = FALSE}
plot(f_MRD_raster)
plot(countriesCoarse, add = T)
```

Let's try changing the colors.

```{r, eval = FALSE}
plot(f_MRD_raster, col = rev(colfuncYellows(100)), axes = FALSE, box = FALSE, 
     zlim = c(minValue(f_MRD_raster), maxValue(f_MRD_raster)), 
     xlab = "Furnariides mean root distance", legend.width = 2)
plot(countriesCoarse, add = T)
```

Let’s plot the two raster.

```{r, eval = FALSE}
par(mfrow = c(1, 2))
plot(f_sr_raster, col = rev(colfuncYellows(100)), axes = FALSE, box = FALSE, 
     zlim = c(minValue(f_sr_raster), maxValue(f_sr_raster)), 
     xlab = "Furnariides richness", legend.width = 2)

plot(f_MRD_raster, col = rev(colfuncYellows(100)), axes = FALSE, box = FALSE, 
     zlim = c(minValue(f_MRD_raster), maxValue(f_MRD_raster)), 
     xlab = "Furnariides mean root distance", legend.width = 2)
```

See if there is a relationship...

```{r, eval = FALSE}
obj <- lm(values(f_MRD_raster) ~ values(f_sr_raster))
cor.test(values(f_sr_raster), values(f_MRD_raster))
```

```{r, eval = FALSE}
plot(values(f_MRD_raster) ~ values(f_sr_raster), xlab = "SR", ylab = "MRD")
abline(obj, col = "red", lwd = 2)
```

Hmmm. What happened in here? So, please answer the next questions.

*From the mean root distance map, it is possible to explain the Furnariides diversity gradient? If so, please explain from an evolutionary perspective*.

## Scale dependency

Now we will explore one of the oldest problems in ecology and evolution, the **scale dependency** in the data. So to explore this scale dependence, we will rasterize the Furnariides ranges, but using different spatial resolutions from 2º to 4º degrees of long-lat.

```{r, eval = FALSE}
# 2º degrees
neo_ras_2dg <- raster()
# Set the raster "extent" 
extent(neo_ras_2dg) <- extent(franges)
res(neo_ras_2dg) <- 2
neo_ras_2dg
values(neo_ras_2dg) <- 0

# 4º degrees
neo_ras_4dg <- raster()
# Set the raster "extent" 
extent(neo_ras_4dg) <- extent(franges)
res(neo_ras_4dg) <- 4
neo_ras_4dg
values(neo_ras_4dg) <- 0
```

```{r, eval = FALSE}
# Furnariides at 2º of long-lat
f_sr_2dg_raster <- rasterize(franges, neo_ras_2dg, field = "SCINAME", fun = function(x,...){length(unique(na.omit(x)))})

# Furnariides at 4º of long-lat
f_sr_4dg_raster <- rasterize(franges, neo_ras_4dg, field = "SCINAME", fun = function(x,...){length(unique(na.omit(x)))})
```

Plot the three maps.

```{r, eval = FALSE}
par(mfrow = c (1, 3))
plot(f_sr_raster, main = "Furnariides richness 1dg")
plot(countriesCoarse, add = T)

plot(f_sr_2dg_raster, main = "Furnariides richness 2dg")
plot(countriesCoarse, add = T)

plot(f_sr_4dg_raster, main = "Furnariides richness 4dg")
plot(countriesCoarse, add = T)
```

So, is there a scale effect?

*Explain the differences between the three maps*

## Correlative relationships

Load the environmental variables that correspond to bio1 (Annual Mean Temperature) and bio12 (Annual Precipitation). These data correspond to two variables out of 19 from WorldClim (http://www.worldclim.org/current). We will use these two variables just for educational purposes, rather to make a complete evaluation of the species-environmental relationships.

```{r, eval = FALSE}
bio1 <- raster("Data/Envi/bio1.bil")
bio1
bio12 <- raster("Data/Envi/bio12.bil")
bio12
```

Plot the environmental variables

```{r, eval = FALSE}
par(mfrow = c(2, 1))
plot(bio1)
plot(bio12)
```

Ok, the bio1 and bio12 layers are at global scale, so now will need to crop to the extent of the Neotropics.

```{r, eval = FALSE}
bio1_neo <- crop(bio1, extent(franges))
bio12_neo <- crop(bio12, extent(franges))
```

```{r, eval = FALSE}
par(mfrow = c(1, 2))
plot(bio1_neo, main = "Annual Mean Temperature")
plot(bio12_neo, main = "Annual Precipitation")
```

Much better!

Now we will obtain the coordinates from the Furnariides diversity raster. These coordinates then will be used to extract the information from the bio1 and bio12 climatic layers.

```{r, eval = FALSE}
f_ras_coords <- xyFromCell(f_sr_raster, 1:length(values(f_sr_raster)))
```

Obtain the values from bio1 and bio12 for each cell or pixel using the coordinates.

```{r, eval = FALSE}
f_ras_bio1 <- extract(bio1_neo, f_ras_coords)
f_ras_bio12 <- extract(bio12_neo, f_ras_coords)

f_ras_rich <- values(f_sr_raster)
f_ras_mrd <- values(f_MRD_raster)
fdata <- na.omit(data.frame(f_ras_coords, f_ras_rich, f_ras_mrd, f_ras_bio1, f_ras_bio12))

#f_ras_rich_noNA <- ifelse(is.na(f_ras_rich), 0, f_ras_rich)
#f_ras_bio1_noNA <- ifelse(is.na(f_ras_bio1), 0, f_ras_bio1)
#f_ras_bio12_noNA <- ifelse(is.na(f_ras_bio12), 0, f_ras_bio12)

#fdata_noNA <- data.frame(f_ras_coords, f_ras_rich_noNA, f_ras_bio1_noNA,
#                         f_ras_bio12_noNA)
```

Now make a simple correlation between the Furnariides richness and bio1 and bio12.

```{r, eval = FALSE}
cor.test(fdata$f_ras_rich, fdata$f_ras_bio1)
```

```{r, eval = FALSE}
cor.test(fdata$f_ras_rich, fdata$f_ras_bio12)
```

*Which environmental variable is more related with Furnariides richness?*

*Please explain the relationship from an ecological perspective*

```{r, eval = FALSE}
par(mfrow = c(1, 2))
plot(fdata$f_ras_bio1, fdata$f_ras_rich, xlab = "Bio 1", ylab = "Richness")

plot(fdata$f_ras_bio12, fdata$f_ras_rich, xlab = "Bio 12", ylab = "Richness")
```

## Considering spatial autocorrelation

A text extracted from **Dorman et al. (2007)**:
The analysis of spatial data is complicated by a phenomenon known as spatial autocorrelation. Spatial autocorrelation **(SAC)** occurs when the values of variables sampled at nearby locations are not independent from each other **(Tobler 1970)**. The causes of spatial autocorrelation are manifold, but three factors are particularly common: 1) biological processes such as speciation, extinction, dispersal or species interactions are distance‐related; 2) non‐linear relationships between environment and species are modelled erroneously as linear; 3) the statistical model fails to account for an important environmental determinant that in itself is spatially structured and thus causes spatial structuring in the response **(Besag 1974)**. Since they also lead to autocorrelated residuals, these are equally problematic. A fourth source of spatial autocorrelation relates to spatial resolution, because coarser grains lead to a spatial smoothing of data. In all of these cases, SAC may confound the analysis of species distribution data.

We now that a correlation is not a causation, so to explore the relationship we need to build a model or fit a model. To explore this relationships we will first explore a simple Ordinary Least Square regression or OLS.

```{r, eval = FALSE}
fols <- lm(f_ras_rich ~ f_ras_bio1 + f_ras_bio12, data = fdata)
summary(fols)
```

Let's complicate our model a little bit...

```{r, eval = FALSE}
fols2 <- lm(f_ras_rich ~ f_ras_bio1 + f_ras_bio12 + f_ras_mrd, data = fdata)
summary(fols2)
```

*What is telling us this OLS?*

Now, explore the spatial autocorrelation of the Furnariides richness gradient.

```{r, eval = FALSE}
autocor_SR <- ncf::correlog(fdata$x, fdata$y, z = fdata$f_ras_rich, na.rm = T, 
                         increment = 1, resamp = 1)
```

Let's use an correlogram to explore the spatial autocorrelation.

```{r, eval = FALSE}
plot(autocor_SR$correlation[1:50], type = "b", pch = 1, cex = 1.2, lwd = 1.5,
     ylim = c(-1, 1), xlab = "Distance class", ylab = "Moran's I", cex.lab = 1.2, 
     cex.axis = 1.2)
abline(h = 0)
```

*Is there a spatial autocorrelation in the data?*

What about the residuals? Now, we will explore the spatial autocorrelation in the residuals.

```{r, eval = FALSE}
coords <- fdata[1:2]
coords <- as.matrix(coords)
```

Build a neighbourhood contiguity by distance

```{r, eval = FALSE}
nb1.5 <- dnearneigh(coords, 0, 1.5)
```

Using the neighbourhood contiguity build a spatial weights for neighbours lists.

```{r, eval = FALSE}
nb1.5.w <- nb2listw(nb1.5, glist = NULL, style = "W", zero.policy = TRUE)
```

Extract the residuals from the OLS model

```{r, eval = FALSE}
residuals_ols <- residuals(fols2)
plot(residuals_ols)
```

Calculate a univariate spatial correlogram.

```{r, eval = FALSE}
autocor_ols_res <- correlog(fdata$x, fdata$y, z = residuals(fols), 
                        increment = 1, resamp = 1)
```

plot the autocorrelagram for the residuals

```{r, eval = FALSE}
plot(autocor_ols_res$correlation[1:50], type = "b", pch = 1, cex = 1.2, lwd = 1.5,
     ylim = c(-0.5, 1), xlab = "distance", ylab = "Moran's I", cex.lab = 1.5, 
     cex.axis = 1.2)
abline(h = 0)
title(main = "OLS residuals", cex = 1.5)
```

Ohhh, seems that the residuals have a strong spatial autocorrelation, that is a problem because if we found autocorrelation in the residuals much of the explanation that we obtain can be biased.

Let's inspect two autocorrelograms.

```{r, eval = FALSE}
par(mfrow = c(2, 1))

plot(autocor_SR$correlation[1:50], type = "b", pch = 1, cex = 1.2, lwd = 1.5,
     ylim = c(-1, 1), xlab = "Distance class", ylab = "Moran's I", cex.lab = 1.2, 
     cex.axis = 1.2)
abline(h = 0)
title(main = "OLS model", cex = 1.5)

plot(autocor_ols_res$correlation[1:50], type = "b", pch = 1, cex = 1.2, lwd = 1.5,
     ylim = c(-0.5, 1), xlab = "Distance class", ylab = "Moran's I", cex.lab = 1.5, 
     cex.axis = 1.2)
abline(h = 0)
title(main = "OLS residuals", cex = 1.5)
```

Hmmm, seems that there is a strong spatial autocorrelation, thus any conclusion using the OLS model can be biased.

To try to solve this important issue, we will use **spatial simultaneous autoregressive error model estimation (Aka SAR model)**, this model account for spatial autocorrelation by adding an extra term (autoregressive) in the form of a spatial-weight matrix that specifies the neighborhood of each cell or pixel and the relative weight of each neighbor.

Let's fit the SAR model.

```{r, eval = FALSE}
sar_nb1.5.w <- errorsarlm(fols2, listw = nb1.5.w, data = fdata, quiet = FALSE, 
                          zero.policy = TRUE, na.action = na.exclude)
summary(sar_nb1.5.w)
residuals_sar_nb1.5.w <- residuals(sar_nb1.5.w)
```

Now estimate the spatial autocorrelation of the SAR model.

```{r, eval = FALSE}
autocor_sar_nb1.5.w <- correlog(fdata$x, fdata$y, z = residuals(sar_nb1.5.w), 
                                na.rm = T, increment = 1, resamp = 1)
```

Plot the autocorrelogram under the SAR model.

```{r, eval = FALSE}
plot(autocor_sar_nb1.5.w$correlation[1:50], type = "b", pch = 4, cex = 1.2, lwd = 1.5,
     ylim = c(-0.5, 1), xlab = "distance", ylab = "Moran's I", cex.lab = 1.5, 
     cex.axis = 1.2)
abline(h = 0)
title(main = "Correlogram SARerr", cex = 1.5)
```

Ohhh, where is the autocorrelation in the residuals? Now compare the two autocorrelograms.

```{r, eval = FALSE}
par(mfrow = c(2, 1))
plot(autocor_ols_res$correlation[1:50], type = "b", pch = 1, cex = 1.2, lwd = 1.5,
     ylim = c(-0.5, 1), xlab = "distance", ylab = "Moran's I", cex.lab = 1.5, 
     cex.axis = 1.2)
abline(h = 0)
title(main = "OLS residuals", cex = 1.5)

plot(autocor_sar_nb1.5.w$correlation[1:50], type = "b", pch = 4, cex = 1.2, lwd = 1.5,
     ylim = c(-0.5, 1), xlab = "distance", ylab = "Moran's I", cex.lab = 1.5, 
     cex.axis = 1.2)
abline(h = 0)
title(main = "Correlogram SARerr", cex = 1.5)
```

Ok, now we know that the SAR model can solve the problem in the spatial autocorrelation in the residuals, let's try to make some inferences.

```{r, eval = FALSE}
summary(sar_nb1.5.w)
```

```{r, eval = FALSE}
summary(fols2)
```

*By looking to the summary of the SAR and OLS models, explain the diferences in the coefficients between both models.*

Now let's compare the prediction of both models. To calculate a R2 to the SAR model, we will use the function **SARr2()** from GitHub.

```{r, eval = FALSE}
source("https://raw.githubusercontent.com/jesusNPL/BetaDivNA/master/SARr2.R")
```

```{r, eval = FALSE}
SARr2(Lfull = sar_nb1.5.w$LL, Lnull = sar_nb1.5.w$logLik_lm.model, N = nrow(fdata))
```

Comparing the two models (OLS and SAR), please answer the following questions:

*1. Which model have the best explanation?*

*2. What can we conclude from these results?*

The end!!! for now...

# References

Besag, J.. 1974. Spatial interaction and the statistical analysis of lattice systems. J. Roy. Stat. Soc. B 36: 192–236.

Brown JH (2012) Why are there so many species in the tropics? J Biogeogr 41:8–22

Currie DJ, Mittelbach GG, Cornell HV et al (2004) Predictions and tests of climate-based hypotheses of broad-scale variation in taxonomic richness. Ecol Lett 7:1121–1134

F. Dormann, C., M. McPherson, J., B. Araújo, M., Bivand, R., Bolliger, J., Carl, G., … Wilson, R. (2007). Methods to account for spatial autocorrelation in the analysis of species distributional data: a review. Ecography, 30(5), 609–628. doi:10.1111/j.2007.0906-7590.05171.x

Fine PVA (2015) Ecological and evolutionary drivers of geographic variation in species diversity. Ann Rev Ecol Evol Syst 46:369–392

Hawkins BA (2001) Ecology’s oldest pattern. Trends Ecol Evol 16:170

Hawkins BA, Field R, Cornell HV et al (2003) Energy, water, and broad-scale geographic patterns of species richness. Ecology 84:3105–3117

Hawkins BA, Diniz-Filho JAF (2004) ‘Latitude’ and geographic patterns in species richness. Ecography 27:268–272

O’Brien EM (2006) Biological relativity to water–energy dynamics. J Biogeogr 33:1868–1888

Pianka ER (1966) Latitudinal gradients in species diversity: a review of concepts. Am Nat 100:33–46

Ricklefs RE (2004) A comprehensive framework for global patterns in biodiversity. Ecol Lett 7:1–15 595

Rohde K (1992) Latitudinal gradients in species diversity: the search for the primary cause. Oikos 65:514–527

Simpson GG (1964) Species densities of North American recent mammals. Syst Zool 13:57–73

Tobler, W. R.. 1970. A computer movie simulating urban growth in the Detroit region. Econ. Geogr. 46: 234–240.
