---
title: 'Brownian Motion'
author: Jesús N. Pinto-Ledezma and Jeannine Cavender-Bares
output:
  pdf_document: default
  html_document: 
    theme: readable
    toc: yes
---

Our understanding of trait evolution over time largely relies on the use of stochastic models for the evolution of continuous traits over phylogenies. The two most widely used models, Brownian motion and the Ornstein–Uhlenbeck (OU) process, differ in that the latter includes constraints on the variance that a trait can attain in a clade.

The OU model explicitly models adaptive evolution toward a trait optimum and has thus been widely used to demonstrate the existence of stabilizing selection on a trait. 

Brownian motion is a stochastic model in which changes from one time to the next are random draws from a normal distribution with mean 0.0 and variance σ2 × Δt. In other words, the expected variance under Brownian motion increases linearly through time with instantaneous rate σ2.

Brownian motion is very easy to simulate. To start off, let's simulate a single instance of Brownian motion for 10, 100 and 1000 generations of discrete time in which the variance of the diffusion process is σ2 = 0.01 per generation.

```{r, warnings = FALSE, message = FALSE}
times <- 0:10  # time
sig2 <- 0.01
## first, simulate a set of random deviates
x <- rnorm(n = length(times) - 1, sd = sqrt(sig2))
## now compute their cumulative sum
x <- c(0, cumsum(x))
plot(times, x, type = "l", ylim = c(-2, 2))
```

```{r, warnings = FALSE, message = FALSE}
times <- 0:100  # time
sig2 <- 0.01
## first, simulate a set of random deviates
x <- rnorm(n = length(times) - 1, sd = sqrt(sig2))
## now compute their cumulative sum
x <- c(0, cumsum(x))
plot(times, x, type = "l", ylim = c(-2, 2))
```

```{r, warnings = FALSE, message = FALSE}
times <- 0:1000  # time
sig2 <- 0.01
## first, simulate a set of random deviates
x <- rnorm(n = length(times) - 1, sd = sqrt(sig2))
## now compute their cumulative sum
x <- c(0, cumsum(x))
plot(times, x, type = "l", ylim = c(-2, 2))
```

No let's repeat the process for a bunch of simulations...

```{r, warnings = FALSE, message = FALSE}
nsim <- 100
X <- matrix(rnorm(n = nsim * (length(t) - 1), sd = sqrt(sig2)), nsim, length(t) - 
    1)
X <- cbind(rep(0, nsim), t(apply(X, 1, cumsum)))
plot(t, X[1, ], xlab = "time", ylab = "phenotype", ylim = c(-2, 2), type = "l")
apply(X[2:nsim, ], 1, function(x, t) lines(t, x), t = t)
```

To see how the outcome depends on σ2, let's compate the result when we divide sig2 by 10:

```{r, warnings = FALSE, message = FALSE}
X <- matrix(rnorm(n = nsim * (length(t) - 1), sd = sqrt(sig2/10)), nsim, length(t) - 
    1)
X <- cbind(rep(0, nsim), t(apply(X, 1, cumsum)))
plot(t, X[1, ], xlab = "time", ylab = "phenotype", ylim = c(-2, 2), type = "l")
apply(X[2:nsim, ], 1, function(x, t) lines(t, x), t = t)
```

There are a number of different ways we could've done this. For example, instead of using apply, which economizes on code, we could have used a for loop as follows:

```{r, warnings = FALSE, message = FALSE}
X <- matrix(0, nsim, length(t))
for (i in 1:nsim) X[i, ] <- c(0, cumsum(rnorm(n = length(t) - 1, sd = sqrt(sig2))))
plot(t, X[1, ], xlab = "time", ylab = "phenotype", ylim = c(-2, 2), type = "l")
for (i in 1:nsim) lines(t, X[i, ])
```

As mentioned above, the expected variance under Brownian motion is just σ2 × t. To see this easiest, we can just do the following. Here I will use 10,000 simulations for 100 generations under the same conditions to "smooth out" our result:

```{r, warnings = FALSE, message = FALSE}
nsim <- 10000
X <- matrix(rnorm(n = nsim * (length(t) - 1), sd = sqrt(sig2)), nsim, length(t) - 
    1)
X <- cbind(rep(0, nsim), t(apply(X, 1, cumsum)))
v <- apply(X, 2, var)
plot(t, v, type = "l", xlab = "time", ylab = "variance among simulations")
```

```{r, warnings = FALSE, message = FALSE}
var(X[, length(t)])  # this should be about 1.00
```

OK, now let's try to simulate using Brownian motion up the branches of a tree. Sticking with discrete time, we first need a discrete time phylogeny, which we can simulate using pbtree in the "phytools" package:


```{r, warnings = FALSE, message = FALSE}
library(phytools)
t <- 100  # total time
n <- 30  # total taxa
b <- (log(n) - log(2))/t
```

```{r, warnings = FALSE, message = FALSE}
tree <- pbtree(b = b, n = n, t = t, type = "discrete")
```

```{r, warnings = FALSE, message = FALSE}
plotTree(tree)
```

Now, to simulate on all the branches of the tree we just need to simulate on each branch separately, and then "shift" each daughter branch by the end state of it's parent. We can do this, remember, because the outcome of Brownian evolution at each time step is independent of all other time steps.

```{r, warnings = FALSE, message = FALSE}
## simulate evolution along each edge
X <- lapply(tree$edge.length, function(x) c(0, cumsum(rnorm(n = x, sd = sqrt(sig2)))))
## reorder the edges of the tree for pre-order traversal
cw <- reorder(tree)
## now simulate on the tree
ll <- tree$edge.length + 1
for (i in 1:nrow(cw$edge)) {
    pp <- which(cw$edge[, 2] == cw$edge[i, 1])
    if (length(pp) > 0) 
        X[[i]] <- X[[i]] + X[[pp]][ll[pp]] else X[[i]] <- X[[i]] + X[[1]][1]
}
## get the starting and ending points of each edge for plotting
H <- nodeHeights(tree)
## plot the simulation
plot(H[1, 1], X[[1]][1], ylim = range(X), xlim = range(H), xlab = "time", ylab = "phenotype")
for (i in 1:length(X)) lines(H[i, 1]:H[i, 2], X[[i]])
## add tip labels if desired
yy <- sapply(1:length(tree$tip.label), function(x, y) which(x == y), y = tree$edge[, 
    2])
yy <- sapply(yy, function(x, y) y[[x]][length(y[[x]])], y = X)
text(x = max(H), y = yy, tree$tip.label)
```

In reality, most simulations of Brownian motion are conducted using continuous rather than discrete time. This is because the additivity of Brownian motion means that the expected variances among & covariances between species are the same in whether we simulate t steps each with variance σ2, or one big step with variance σ2t.

Here is an example of a continuous time simulation & visualization using canned functions.

```{r, warnings = FALSE, message = FALSE}
## simulate Brownian evolution on a tree with fastBM
x <- fastBM(tree, sig2 = sig2, internal = TRUE)
## visualize Brownian evolution on a tree
phenogram(tree, x, spread.labels = TRUE, spread.cost = c(1, 0))
```

Some additional points

A few additional points about Brownian evolution were made by Joe in his lecture on the topic, so I have added some additional simulations to illustrate these points.

Joe pointed out that in some cases there might be different rates of Brownian evolution in different parts of the tree. It is straightforward to simulate Brownian motion with different rates on different branches. For example, in the following example, I first simulate the history of a discretely valued character on the tree, and then a continuous trait with a rate that changes as a function of the mapped discrete character. I'll then create a traitgram that projects the phylogeny onto the phenotypic trait axis and overlays the mapped discrete character.

```{r, warnings = FALSE, message = FALSE}
set.seed(100)
## transition matrix for the discrete trait simulation
Q <- matrix(c(-1, 1, 1, -1), 2, 2)
## simulated tree
tree <- pbtree(n = 30, scale = 1)
## simulate discrete character history
tree <- sim.history(tree, Q, anc = "1")
## plot discrete character history on the tree
plotSimmap(tree, setNames(c("blue", "red"), 1:2), pts = F)
```

```{r, warnings = FALSE, message = FALSE}
x <- sim.rates(tree, setNames(c(1, 20), 1:2), internal = TRUE)
phenogram(tree, x, colors = setNames(c("blue", "red"), 1:2), spread.labels = TRUE, 
    spread.cost = c(1, 0))
```

Joe discussed the fact that shared ancestry creates "covariance" among the tips of the tree. In my presentation, I described how this is equivalent to covariance among species across a large number of replications of the evolutionary process. The following is an illustration of that point. (This requires the package "car".)

```{r, warnings = FALSE, message = FALSE}
## simulate 5 taxon tree
tree <- pbtree(n = 5)
## 500 BM simulations
X <- fastBM(tree, nsim = 500)
## plot tree
plot(tree, edge.width = 2, direction = "downwards")
```

```{r, warnings = FALSE, message = FALSE}
## plot distribution across tips from 500 simulations
require(car)
scatterplotMatrix(t(X))
```

Joe pointed out that Brownian motion does not assume that the process under which the individual lineages are moving is Gaussian. The result will be Gaussian - due to the Central Limit Theorem. Here is an illustration of that:

```{r, warnings = FALSE, message = FALSE}
t <- 0:100  # time
sig2 <- 0.01
nsim <- 1000
## we'll simulate the steps from a uniform distribution with limits set to
## have the same variance (0.01) as before
X <- matrix(runif(n = nsim * (length(t) - 1), min = -sqrt(3 * sig2), max = sqrt(3 * 
    sig2)), nsim, length(t) - 1)
X <- cbind(rep(0, nsim), t(apply(X, 1, cumsum)))
plot(t, X[1, ], xlab = "time", ylab = "phenotype", ylim = c(-2, 2), type = "l")
apply(X[2:nsim, ], 1, function(x, t) lines(t, x), t = t)
```

```{r, warnings = FALSE, message = FALSE}
var(X[, length(t)])
```

```{r, warnings = FALSE, message = FALSE}
hist(X[, length(t)])
```

```{r, warnings = FALSE, message = FALSE}
plot(density(X[, length(t)]))
```

Thomas pointed out that Brownian motion is generally assumed to be trendless; however it is possible to simulate and (under some conditions) fit a model of Brownian evolution with a trend. Here is an example of a simulation (using the same general approach as above) with a trend.

```{r, warnings = FALSE, message = FALSE}
nsim = 100
X <- matrix(rnorm(mean = 0.02, n = nsim * (length(t) - 1), sd = sqrt(sig2/4)), 
    nsim, length(t) - 1)
X <- cbind(rep(0, nsim), t(apply(X, 1, cumsum)))
plot(t, X[1, ], xlab = "time", ylab = "phenotype", ylim = c(-1, 3), type = "l")
apply(X[2:nsim, ], 1, function(x, t) lines(t, x), t = t)
```

Finally, Joe mentioned the threshold model - so I'm obligated to show that "phytools" can do simulation and plotting under the threshold model. Here is an example:

```{r, warnings = FALSE, message = FALSE}
tree <- pbtree(n = 100, scale = 1)
xx <- bmPlot(tree, sig2 = 2/1000, type = "threshold", thresholds = c(-2, 0, 2))
```



