---
title: "Ecological niche models and species distributions"
description: Showing some neat features of R!
date: today # other options: now, last-modified
authors:
  - name: Jesús N. Pinto-Ledezma 
    url: https://jesusnpl.github.io
    affiliation: Ecology, Evolution & Behavior, University of Minnesota
                Biodiversity Science - EBB 3534-5534
    affiliation-url: https://github.com/jesusNPL/BiodiversityScience
title-block-banner: true
#title-block-banner: "#562457"
format: 
  html: 
    embed-resources: true # save contained file.
    theme: spacelab # check other themes https://quarto.org/docs/output-formats/html-themes.html 
    code-fold: true
    toc: true
    number-sections: true
bibliography: lab_5_ENM-SDM.bib
link-citations: true
csl: ../apa-single-spaced.csl
---

:::{.callout-note}

In this lab, we will explore some Correlative models using data at species level (occurrence data points) and environmental data at large spatial scales. We will download occurrence data for oak species distributed in the Americas from [**GBIF**](https://www.gbif.org) and environmental (climatic) data from [**WorldClim**](https://www.worldclim.org).

:::

# Set up your data and your working directory

To do this laboratory you will need to install the following packages:

```{r}
#| eval: true

packages <- c("terra", "geodata", "sp", "spThin", "sf", "maps", "dismo", 
              "CoordinateCleaner", "rgbif", "countrycode", "rworldmap", 
              "Metrics", "ROCR", "ggraph", "plotly")
# Package vector names

```

:::{.callout-tip}
## Function install.packages()
You can use the function **install.packages()** to install the packages.
:::

If you don't want to install the packages one by one, you can use the next command. 

```{r}
#| eval: true

# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())

if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages], dependencies = TRUE)
}

```

Although we installed several packages in the previous lines two packages are missing, {**torch**} and {**cito**}. These two  packages are necessary for modeling the environmental niches (**shadows**) or Grinnellian niches under a **deep learning** framework--in other words, its a method that teaches computers to process data in a way that is inspired by the human brain. {**torch**} is the main package and {**cito**} is a wrapper that facilitate the implementation of **deep neural networks**. We will install these packages separately because.

Let's start with {**torch**}

```{r}
#| eval: true

# check package 
if(!require('torch', quietly = TRUE)) install.packages('torch')
library('torch') 

#install torch
if(!torch_is_installed()) install_torch()

```

A last step is to install {**cito**}

```{r}
#| eval: true

if (!("cito" %in% installed.packages())) {
    install.packages("cito")
}

```

Call the packages

```{r}
#| eval: true

sapply(packages, require, character.only = TRUE)

library(tidyverse)

```

Double-check your working directory. 

:::{.callout-tip}
## Function getwd()
You can use the function **getwd()** to get the current working directory.
:::

# Prepare data

## Download species occurrences from GBIF

Get some occurrence data for our species from GBIF, directly within R. This may take some time, given the number of occurrences for the selected species. 

**NOTE**: we need to have an internet connection.

Set the vector with some species scientific names (N = 5) to download occurrence data from **GBIF**.

```{r}
#| eval: true

spp <- c("Quercus virginiana", "Quercus minima", "Quercus alba", "Quercus fusiformis", "Quercus oleoides")

```

Download data using the vector of species names

```{r}
#| eval: true

gbif_data <- occ_search(scientificName = spp, 
                      hasCoordinate = TRUE, 
                      limit = 2000)  

#decrease the 'limit' if you just want to see how many records there are without waiting all the time that it will take to download the whole dataset.

```

Save the raw data

```{r}
#| eval: false

dir.create("data")
dir.create("data/OCC")

# save the donwloaded in the working directory
save(gbif_data, file = "data/OCC/oaks_raw_occ.RDATA")

```

Inspect the downloaded data. Note that we have downloaded data for five species but for simplicity let's pick one species for this lab. I will select *Quercus minima* but you can select any other species.

The returned object is a list of five elements (i.e., the number of our species vector), thus, to select the focal species just use **[[position of the species in the vector]]** to get access to the species. For example, I selected *Quercus minima*, so, to get access to this species I just need to type **gbif_data[[2]]**, let's see:

```{r}
#| eval: false

#gbif_data
# if, for any species, "Records found" is larger than "Records returned", you need to increase the 'limit' argument above -- see help(occ_data) for options and limitations

# check how the data are organized:
names(gbif_data)

# metadata of the object #2
names(gbif_data[[2]]$meta)

# the data
names(gbif_data[[2]]$data)

```

Wow, there are a bunch of columns, let's select some columns that are relevant for our purposes.

```{r}
#| eval: true

# get the columns that matter for mapping and cleaning the occurrence data:
occ_quemin <- gbif_data[[2]]$data %>% 
  dplyr::select(species, decimalLongitude, 
                decimalLatitude, countryCode, individualCount,
                gbifID, family, taxonRank, coordinateUncertaintyInMeters,
                year, basisOfRecord, institutionCode, datasetName)

occ_quemin

```

Let's look the data.

```{r}
#| eval: true

glimpse(occ_quemin)

```

Now let's clean the data a little bit. First, removing all NA in the occurrence data.

```{r}
#| eval: true

occ_quemin <- occ_quemin %>% 
  filter(!is.na(decimalLongitude & !is.na(decimalLatitude)))

occ_quemin

```

Cleaning geographic coordinates by flagging potentially erroneous locations.

```{r}
#| eval: true

#convert country code from ISO2c to ISO3c
occ_quemin$countryCode <- countrycode::countrycode(occ_quemin$countryCode, 
                                                   origin = 'iso2c', 
                                                   destination = 'iso3c')

#flag problems
occ_quemin <- data.frame(occ_quemin)

occ_quemin_flag <- clean_coordinates(x = occ_quemin, 
                           lon = "decimalLongitude", 
                           lat = "decimalLatitude",
                           countries = "countryCode",
                           species = "species",
                           tests = c("capitals", "centroids",
                                    "equal", "zeros", "countries")) # most test are on by default

```
See the results 

```{r}
#| eval: true

summary(occ_quemin_flag)

```

It seems that 23 observations were flagged, let's plot the points to see the which ones.

```{r}
#| eval: true

plot(occ_quemin_flag, lon = "decimalLongitude", lat = "decimalLatitude")

```

Exclude or filter the flagged records 

```{r}
#| eval: true

occ_quemin_clean <- occ_quemin_flag %>% 
  filter(.summary == TRUE)

#occ_quemin_clean

```

Plot the cleaned data 

```{r}
#| eval: true

plot(occ_quemin_clean, lon = "decimalLongitude", lat = "decimalLatitude")

```

Now transform our occurrence data to a spatial object, this will help us to visualize better our data.

```{r} 
#| eval: true

quemin_spt <- st_as_sf(occ_quemin_clean, 
                       coords = c("decimalLongitude", "decimalLatitude"), 
                       crs = 4326) #%>%
  #st_cast("MULTIPOINT")

# explore the data
glimpse(quemin_spt)

```

Let's plot the distribution of *Quercus minima* occurrences.

```{r}
#| eval: true

sf_use_s2(FALSE)

# world map
worldMap <- rnaturalearth::ne_countries(scale = "medium", 
                                        type = "countries", 
                                        returnclass = "sf")

# country subset
USpoly <- worldMap %>% 
  #filter(region_wb == "North America")
  filter(admin == "United States of America")

```

Plot the occurrence records

```{r}
#| eval: true

map_US <- map_data('world')[map_data('world')$region == "USA", ]

ggplot() + 
  geom_polygon(data = map_data("world"), 
               aes(x = long, y = lat, group = group), 
               color = "#f1f2f3", fill = '#f3f3f3') + 
  geom_polygon(data = map_US, 
               aes(x = long, y = lat, group = group), 
               color = 'lightgray', fill = 'lightgray', alpha = 0.5) + 
  geom_point(data = occ_quemin, 
             aes(x = decimalLongitude, y = decimalLatitude), 
             color = "darkgray", alpha = 0.9) + 
  coord_map() + 
  coord_fixed(1.3, 
              xlim = c(-65, -125), 
              ylim = c(50, 25)
              ) + 
  theme(panel.background = element_rect(fill = "lightblue"))

```

Small joke 😬🫣. 

Now the real map...

```{r}
#| eval: true

ggplot() + 
  geom_polygon(data = map_data("world"), 
               aes(x = long, y = lat, group = group), 
               color = "#f1f2f3", fill = '#f3f3f3', alpha = 0.5) + 
  geom_sf(data = USpoly) + 
  geom_sf(data = quemin_spt, color = "darkgray", alpha = 0.9) + 
  coord_sf(
     xlim = c(-125, -65), 
     ylim = c(25, 50)
  ) + 
  theme(
    plot.background = element_rect(fill = "#f1f2f3"),
    panel.background = element_rect(fill = "lightblue"),
    panel.grid = element_blank(),
    line = element_blank(),
    rect = element_blank()
  )

```

Seems the data is correct but sometimes the data aggregation can cause some troubles in model predictions. To solve this issue we will apply a spatial thinning on the species occurrences, to do so, we will use the package {**spThin**}. Given that our species has a broad distribution let's set a distance threshold of 2 km between occurrence points.

```{r}
#| eval: true

quemin_thinned <- thin(
      loc.data =  occ_quemin_clean, 
      verbose = FALSE, 
      long.col = "decimalLongitude", 
      lat.col = "decimalLatitude",
      spec.col = "species",
      thin.par = 2, # points have at least a minimum distance of 2 km from each other
      reps = 10, # number of repetitions
      locs.thinned.list.return = TRUE, 
      write.files = FALSE, 
      out.dir = "data/OCC/")
    
quemin_thinned <- as.data.frame(quemin_thinned[[10]])
quemin_thinned$species <- "Quercus_minima"

```

Explore the results of the thinning process. We can see that the number of occurrences of *Quercus minima* decreased from 784 occurrences to 440 occurrences. We will use the thinned data for further analyses.

```{r}
#| eval: true

glimpse(quemin_thinned)

```

Transform the thinned occurrences to a spatial object

```{r}
#| eval: true

quemin_thinned_sf <- st_as_sf(quemin_thinned, 
                       coords = c("Longitude", "Latitude"), 
                       crs = 4326) 

# The below code is using R base. We will use this object for the modelling part
quemin_thinned_spt <- SpatialPointsDataFrame(coords = quemin_thinned[, 1:2], 
                                             data = quemin_thinned, 
                                             proj4string = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")) 


```

Now visualize the both spatial objects.

```{r}
#| eval: true

ggplot() + 
  geom_polygon(data = map_data("world"), 
               aes(x = long, y = lat, group = group), 
               color = "#f1f2f3", fill = '#f3f3f3') + 
  geom_sf(data = USpoly) + 
  geom_sf(data = quemin_spt, color = "blue", alpha = 0.5) + 
  geom_sf(data = quemin_thinned_sf, color = "red", alpha = 0.3) + 
  coord_sf(
     xlim = c(-125, -65), 
     ylim = c(25, 50)
  ) +
  theme(
    plot.background = element_rect(fill = "#f1f2f3"),
    panel.background = element_rect(fill = "lightblue"),
    panel.grid = element_blank(),
    line = element_blank(),
    rect = element_blank()
  )

```

Now let's save the data processed and keep **thinned_spt** for further analyses.

```{r}
#| eval: true

save(occ_quemin, quemin_spt, quemin_thinned_sf, quemin_thinned, 
     file = "data/OCC/quemin_OCC_processed.RData")

rm(gbif_data, occ_quemin, quemin_spt)

```

Until here we have downloaded and cleaned species occurrence data, now we will prepare the environmental data that is going to be used as predictors for constructing ENM/SDM for our species.

## Prepare environmental data

First we will obtain bioclimatic variables from **WorldClim** and to do that we will use the function '**worldclim_global()**' from the package {**geodata**}, then we will select which variables are more relevant for explaining the distribution of our species.

```{r}
#| eval: false

bios <- geodata::worldclim_global(var = "bio", # climatic variables
                         res = 10, # spatial resolution
                         path = "data/", # folder location
                         version = "2.1")


```

You can see that by using the argument **path** within the function **worldclim_global()** the bioclimatic variables were also downloaded directly to your hard drive, so you don't need to download the data again. 

:::{.callout-note}

## Download the climatic data manually
If you have issues downloading WorldClim data, you can download the data [HERE](https://www.dropbox.com/scl/fi/iwvoys8t0hj90vb3hrz0a/wc2.1_10m.zip?rlkey=86m5wrvxwaci5fhskwffqrhjj&dl=0 ) and then store the climatic data within the folder **data**. 

:::

### Load bioclimatic data stored in your computer

If you downloaded the bioclimatic data from dropbox please use the next lines of code.

```{r}
#| eval: true

biosNames <- list.files("data/wc2.1_10m/", 
                        pattern = "tif$")

# Please double check your folder path. 

bios <- rast(paste0("data/wc2.1_10m/", # folder directory 
                     biosNames)) # file names

```


Let's explore some details of the bioclimatic data 

```{r}
#| eval: true

names(bios)

str(bios[[1]])

```

Now let's plot some bioclimatic variables, let's say BIO1 and BIO12 that are mean annual temperature and annual precipitation, respectively.

```{r}
#| eval: true

plot(c(bios[[1]], bios[[12]]))

```
Okay, the environmental data is at global level but our species only occur in North America, in order to have a better visualization of the data, let's crop the bioclimatic data to the extent of the United States. 

```{r}
#| eval: true

USpoly_spt <- as(USpoly, "Spatial") # used as extent

plot(USpoly_spt)

```

Now crop the bioclimatic data to the extent of North America.

```{r}
#| eval: true

bios_US <- crop(bios, USpoly, mask = TRUE)

```

Let's visualize the results of cropping the bioclimatic data and overlap the thinned occurrences of *Quercus minima*.

```{r}
#| eval: true

plot(bios_US[[1]]) # mean annual temperature
plot(quemin_thinned_spt, col = "red", pch = 16, add = TRUE) # add occurrence records
plot(USpoly_spt, lwd = 2, lty = 2, add = TRUE) # add country borders

```
Cool!

# Accessible area

Before any further analysis we need to define the accessible area for our species (*Quercus minima*) in the geographical space, remember our species is distributed in the United States. To define the accessible area let's use a bounding box around the known species occurrences plus ~300 km beyond that bound. This will give us an approximation of the species dispersal within the geographical domain, i.e., North America. This bounding box represent the component **M** in the **BAM** diagram.

```{r}
#| eval: true

### Species specific accessible area
bb <- bbox(quemin_thinned_spt) # bounding box

ex <- ext(c(bb[1]-3, bb[3]+3, bb[2]-3, bb[4]+3)) # bounding box + 300 km

pex <- as.polygons(ex, crs = crs(bios_US)) #as(ex, 'SpatialPolygons') # transform to polygon
#crs(pex) <- crs(bios_US) # use the geographical reference of the bioclimatic variables

crs(USpoly_spt) <- crs(bios_US)

out <- terra::crop(USpoly_spt, as(pex, 'Spatial'), byid = FALSE) # use NAs to eliminate areas on the sea

```

Now let's plot the accessible area.

```{r}
#| eval: true

plot(bios_US[[1]])
plot(pex, add = TRUE, lty = 2)
plot(out, add = TRUE, lwd = 2)

```

Nice, we can see that the accessible area for our species is the southeast United States.

Now crop the bioclimatic data to the extent of the accessible area.

```{r}
#| eval: true

bios_spp <- terra::crop(bios, out, mask = TRUE)
#bios_spp <- terra::mask(bios_spp, vect(out))

# plot the results
plot(bios_spp[[1]])
#plot(quemin_thinned_spt, add = TRUE, col = "red", pch = 16)
plot(USpoly_spt, add = TRUE, lty = 2)

```

Note that we will use this accessible area as extent for all the posterior analyses.

# Pseudoabsences

To model the environmental niches (**ghosts**) and to project the species distributions (**shadows**) we need to have data of species presences and absences, however we only have presence data. What should we do if we don't have absences? That's the question! There is no a straightforward answer for that question, but the most common procedure to get absence data is to generate random points (AKA **pseudoabsences**) on the geographical domain (i.e., the accessible area or M). There is a lot of discussion on how many pseudoabsences we need to use for ENM/SDM, here we will use a conservative number, i.e., twice the number presences.

```{r}
#| eval: true

set.seed(12345) # Random Number Generation to obtain the same result

# Generate the data
absence <- dismo::randomPoints(mask = as(bios_spp[[1]], "Raster"), # transform spatraster to raster
                        n = round(nrow(quemin_thinned)*2, 0), # number of pseudoabsences
                        p = quemin_thinned_spt, # presence object
                        ext = c(bb[1]-3, bb[3]+3, bb[2]-3, bb[4]+3) # extent
                        )

```

Now let's combine the **presence** and **pseudoabsence** data

```{r}
#| eval: true

# Presences
presence <- data.frame(quemin_thinned) # presence data

# Pseudoabsences
absence <- data.frame(absence) # pseudoabsence data
absence$species <- "Quercus_minima"
names(absence) <- names(presence)

presence$Occurrence <- 1 # presence data
absence$Occurrence <- 0 # pseudoabsence data

quemin <- rbind(presence, absence) # combine both information

```

Finally transform the **presence-pseudoabsence** data to a spatial object and visualize it!

```{r}
#| eval: true

quemin_sp <- quemin

coordinates(quemin_sp) <- ~ Longitude + Latitude
crs(quemin_sp) <- crs(bios_spp)

quemin_sp

quemin_sf <- st_as_sf(quemin, 
                       coords = c("Longitude", "Latitude"), 
                       crs = crs(bios_spp)) 


```

Let's visualize the results

```{r}
#| eval: true

plot(bios_spp[[1]])
plot(quemin_sp[quemin_sp$Occurrence == 1, ], col = "red", add = TRUE, pch = 16) # presence
points(quemin_sp[quemin_sp$Occurrence == 0, ], col = "blue", pch = 16) # absence

```

We can save the processed data and clean the environment a little bit.

```{r}
#| eval: true

save(presence, absence, quemin, file = "data/OCC/quemin_PresAbs.RData")

save(bb, ex, USpoly_spt, out, pex, file = "data/accessible_area_quemin.RData")

rm(absence, presence, bios, ex, out, pex, bb)

```

Now, we need to decide which variables are necessary to model the niche (Grinnellian niche or Fundamental niche or the Ghost) for our oak species. Here we can rely the selection of the variables by asking a specialist or use statistical tools or both. 

# Variable selection

Let's use the statistical way. 

```{r}
#| eval: true

## extract variables
quemin_bios <- data.frame(terra::extract(bios_spp, quemin_sf))

## Combine coordinates and climate data
quemin_bios <- cbind(data.frame(quemin), quemin_bios)

## Remove NAs
quemin_bios <- quemin_bios[complete.cases(quemin_bios), ]
quemin_bios <- na.omit(quemin_bios)

glimpse(quemin_bios)

```

One way to select variables is to explore the correlation among the predictors. We can use a threshold of **|r| < 0.7** which means that correlated variables below this threshold can be considered as no problematic [@dormann_collinearity_2013], however you can use a more conservative threshold, such as <0.5. 

To do that, we first estimate a correlation matrix from the predictors. We use Spearman rank correlation coefficient, as we do not know whether all variables are normally distributed.

```{r}
#| eval: true

cor_mat <- cor(quemin_bios[, c(6:24)], method = 'spearman')

```

Let's visualize the correlation matrix.

```{r}
#| eval: true

corrplot::corrplot.mixed(cor_mat, 
                         tl.pos = "lt", 
                         tl.cex = 0.5, 
                         number.cex = 0.5, 
                         addCoefasPercent = TRUE, 
                         mar = c(0, 0, 1, 0))

```

Looks like that several variables are highly correlated, but let's select the ones that are highly correlated with each other and exclude them for further analysis. 

Inspecting predictor by predictor can take forever, **what should we do?** Well we have two options:

1. Talk with the specialist about which predictors should we use, or 

2. Let the computer decide 

The simple solution is letting the computer decide, thus, for doing that, we will use an amazing function called '**select07()**' from the package {**mecofun**} that will select the predictors bellow the correlation threshold.

I extracted the function **select07()** from the package {**mecofun**} so we can use it directly without install the {**mecofun**}.

```{r}
#| eval: true

set.seed(12345)

#source("R-Functions/select07_mod.R")
source("https://raw.githubusercontent.com/jesusNPL/BiodiversityScience/master/Spring2024/Lab_5_ENM_SDM/functions/select07_mod.R")

# Run select07()
covar_sel <- select07_v2(X = quemin_bios[, -c(1:5)], # only predictors data 
                         y = quemin_bios$Occurrence, # presence-absence data 
                         threshold = 0.7) # here you can change the threshold for one 

```

Note that if you want to install {**mecofun**} you can use the code below. Important, a message will appear in your console asking if you would like to update an existing package, please type **3** and press **Enter**. This will order **R** to install just {**mecofun**} and also will avoid any error in the installation.

```{r}
#| eval: false

# Install the package from source
remotes::install_git("https://gitup.uni-potsdam.de/macroecology/mecofun.git")

```

The function will return a list of three objects:

1. AIC values for each model

2. The correlation matrix

3. A vector of the selected variables.

```{r}
#| eval: true

# Check out the structure of the resulting object:
str(covar_sel)

```

Let's see the results...

```{r}
#| eval: true

covar_sel$AIC
covar_sel$cor_mat
covar_sel$pred_sel

```

According to the **select07()** function eight climatic variables (predictors) best fit the data and also have low correlation. Assuming that this is correct, we will use these variables for further analyses. The selected variables are: "bio18", "bio4", "bio9", "bio5", "bio15", "bio2", "bio19". Store these variable names as a vector.

```{r}
#| eval: true

preds <- covar_sel$pred_sel
preds

```

Finally, let's select the bioclimatic variables and plot them. Note that we will use these environmental layers for model predictions.

```{r}
#| eval: true

bios_quemin_sel <- bios_spp[[preds]]

plot(bios_quemin_sel)

```

As a pre-final step we will scale the predictors 

```{r}
#| eval: true

quemin_bios_sel <- quemin_bios %>% 
  select(Occurrence, all_of(preds))

quemin_bios_sel[, 2:8] <- scale(quemin_bios_sel[, 2:8])

```

Okay, we are ready to build the models for our species.

# Building Environmental Niche Models

To model the environmental niches for *Quercus minima* we will use the R package {**cito**}. The first thing we need to do is to filter some data for training and some data for testing. We will use ~30% of the data for testing and ~70% for training.

```{r}
#| eval: true

set.seed(12345)

index <- sample.int(nrow(quemin_bios_sel), 400) # 400 points or 0.33% of the data 

## Testing data
testing <- quemin_bios_sel[index, ] 

## Data for training our model
training <- quemin_bios_sel[-index, ]

```

Now let's model the species-environment relationship or Grinnellian niches. 

```{r}
#| eval: true

library(cito)

```

Run Deep Neural Network

```{r}
#| eval: true

set.seed(12345)

quemin_DNN <- dnn(Occurrence ~., # formula
            data = training, # data for training
            batchsize = 150L, 
            epochs = 100L,
            lr = 0.05, 
            #lambda = 0.001,
            #alpha = 1.0,
            validation = 0.33, 
            loss = "binomial", # model 
            #early_stopping = 15, 
            bootstrap = 20L, # number of bootstraps set to 20
            verbose = FALSE)

```

Visualize Network

```{r}
#| eval: false

plot(quemin_DNN)

```

Visualize the bootstrap output. If an error is returned, you can skip this part.

```{r}
#| eval: true

analyze_training(quemin_DNN)

```

Until here we have downloaded and processed occurrence data and fitted a deep learning algorithm. Now let's predict the Grinnellian niche for *Quercus minima* in the accessible area. But first let's explore how good our model is. In addition, you can use the fitted object (i.e., quemin_DNN) to make predictions to other regions (e.g., areas with similar environmental conditions or invadable areas) or to project to other periods of time (i.e., to the future or the past).

```{r}
#| eval: true

quemin_DNN_pred <- predict(quemin_DNN, 
                      newdata = testing, 
                      reduce = "none")

dim(quemin_DNN_pred)

```

Let's get the AUC (Area under the ROC Curve) from the prediction object.

```{r}
#| eval: true

AUCs <- sapply(1:20, function(i) Metrics::auc(testing$Occurrence, quemin_DNN_pred[i,,]))

hist(AUCs,
     xlim = c(0.7, 1), 
     main = "AUC of bootstrap model", 
     xlab = "AUC")

```

We can see that the AUC is very high ranging between 0.980 and 0.984.

Get variable (feature) importance or the variables that are more important in predicting the Grinnellian niche of our species.

```{r}
#| eval: true

quemin_Response <- summary(quemin_DNN)
quemin_Response

```

*How do you feel about that?*

*Which variable is, in fact, not that important in predicting the Grinnellian niche of Quercus minima?* Hint: look at the Feature Importance section in the object **quemin_Response**

## Biological inferences

We can also go a step ahead and make some inferences. For example, we can see how the species is responding to the set of environmental variables we use to model its Grinnellian niche.

```{r}
#| eval: true

ALE(model = quemin_DNN, variable = preds[1])
ALE(model = quemin_DNN, variable = preds[2])
ALE(model = quemin_DNN, variable = preds[3])
ALE(model = quemin_DNN, variable = preds[4])
ALE(model = quemin_DNN, variable = preds[5])
ALE(model = quemin_DNN, variable = preds[6])
ALE(model = quemin_DNN, variable = preds[7])

```

*How do you feel about that?*

Now let's predict the Grinnellian niche for *Quercus minima* in the accessible area.

Auxiliary functions to perform predictions obtained from  (https://citoverse.github.io/cito/articles/C-Example_Species_distribution_modeling.html)

```{r}
#| eval: true

## mean prediction
customPredictFun <- function(model, data) {
  return(apply(predict(model, data, reduce = "none"), 2:3, mean)[, 1])
}

## standar deviation or uncertainty associated to the prediction
customPredictFun_sd <- function(model, data) {
  return(apply(predict(model, data, reduce = "none"), 2:3, sd)[, 1])
} 

```

Perform model prediction

```{r}
#| eval: true

## Scale the predictors 
bios_quemin_sel_scale <- terra::scale(bios_quemin_sel)

## Predict the Grinnellian niche for Quercus minima
quemin_ENM <- terra::predict(object = bios_quemin_sel_scale,
                  model = quemin_DNN,
                  fun = customPredictFun, 
                  na.rm = TRUE)

```

Let’s try making a nice map.

```{r}
#| eval: true

library(rasterVis)
library(RColorBrewer)

mapTheme <- rasterTheme(region = rev(brewer.pal(11, "Spectral")),
  layout.widths = list(right.padding = 10),
  axis.line = list(col = "transparent"),
  tick = list(col = 'transparent'))

levelplot(quemin_ENM,
  maxpixels = 1e10,
  margin = FALSE,
  par.settings = mapTheme,
  scales = list(x = list(draw = TRUE),
                y = list(draw = TRUE)),
  zlim = c(0, 1))

```

We can also predict the uncertainty of our model

```{r}
#| eval: true

quemin_ENM_uncertainty <- terra::predict(object = bios_quemin_sel_scale,
                  model = quemin_DNN,
                  fun = customPredictFun_sd, 
                  na.rm = TRUE)

```

Plot the uncertainty in our prediction

```{r}
#| eval: true

levelplot(quemin_ENM_uncertainty,
  maxpixels = 1e10,
  margin = FALSE,
  par.settings = mapTheme,
  scales = list(x = list(draw = TRUE),
                y = list(draw = TRUE)),
  zlim = c(0, 1))

```

You might also want to save the predictions.

```{r}
#| eval: true

dir.create("results")
dir.create("results/ENM")

# save the predictions
writeRaster(quemin_ENM, 
            filename = "results/ENM/quemin_ENM.tif", 
            overwrite = TRUE)

# save the uncertainty prediction
writeRaster(quemin_ENM_uncertainty, 
            filename = "results/ENM/quemin_ENM_uncertainty.tif", 
            overwrite = TRUE)

```

Finally, let's produce "shadows" from the "ghosts".

# Building Species Distributions

## Species geographical distributions

First we need to set a binarization threshold for the model predictions, the threshold (cut-off) is used to transform model predictions (probabilities, distances, or similar values) to a binary score (presence or absence). 

```{r}
#| eval: true

## Get mean and sd predictions
meanPred <- apply(as.data.frame.array(quemin_DNN_pred), MARGIN = 2, FUN = mean)
sdPred <- apply(as.data.frame.array(quemin_DNN_pred), MARGIN = 2, FUN = sd)

## Create a prediction object
predOBJ <- ROCR::prediction(meanPred, testing$Occurrence)

## Get global AUC
quemin_AUC <- performance(predOBJ, "auc")@y.values[[1]]

## Plot AUC
ROCdf <- performance(predOBJ, "tpr", "fpr")

ROCdf <- data.frame(fpr = ROCdf@x.values[[1]], 
                    tpr = ROCdf@y.values[[1]])

ROCdf %>% 
  ggplot(aes(x = fpr, y = tpr)) + 
  geom_line(linewidth = 2, color = "darkgray") + 
  geom_abline(intercept = 0, slope = 1, color = 'red', linewidth = 2) + 
  ggtitle('Receiver-operator curve') + 
  xlab('False positive rate') + 
  ylab('True positive rate') + 
  theme_classic()

```

Here we will find the threshold for the **mean** prediction, but you can find the threshold for each bootstrap. The threshold we will use intents to maximizes the **sensitivity** (ability of a test to correctly identify presences or true positive rate) plus **specificity** (ability of a test to correctly identify absences or true negative rate) in the predictions.

```{r}
#| eval: true

## Get sensitivity and specificity
quemin_PERF <- ROCR::performance(predOBJ, "sens", "spec") 

PERF_list <- (quemin_PERF@x.values[[1]] + quemin_PERF@y.values[[1]] - 1)

PERF_df <- data.frame(alpha = quemin_PERF@alpha.values[[1]], tss = PERF_list)

## Calculate Threshold
quemin_threshold <- min(PERF_df$alpha[which(PERF_df$tss == max(PERF_df$tss))])

## Calculate True skill statistics (TSS) 
quemin_TSS <- PERF_df[which(PERF_df$alpha == quemin_threshold), 'tss'] 

```

The threshold that maximizes the sensitivity and specificity is **0.39**, we will use that value to produce "shadows" from the "ghosts".

```{r}
#| eval: true

quemin_SDM <- quemin_ENM

# reclassify our ensemble prediction
quemin_SDM[] <- ifelse(quemin_ENM[] >= quemin_threshold, 1, 0)

```

Plot the distribution of *Quercus minima*

```{r}
#| eval: true

plot(quemin_SDM)
plot(countriesCoarse, add = TRUE)

```

That's it, we modeled environmental niches (**ghosts**) to produce geographical distributions (**shadows**).

You might also want to save this final raster file in your hard drive.

```{r}
#| eval: true

dir.create("results/SDM")

writeRaster(quemin_SDM, 
            filename = "results/SDM/quemin_SDM.tif", 
            overwrite = TRUE)

```

# The challenge

Please respond each question based on the practice.

1. Do you think the observations are a reasonable representation of the distribution (and ecological niche) of the species? Explain your answer.

2. Can we use ENM/SDM to aid the conservation of *Quercus minima*? Explain your answer.

3. Based on the lecture and the practice, what can we conclude?

4. Model the niche and predict the distribution of a different species (you can use one of the species listed at the beginning of the class).

The end, for now 😉!